Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-25 manu-draft.Rmd
minor edits to wq results, wqmap scale updated for tn

The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.  Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.  Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.  Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater from the stack to be released into Bishop Harbor, with an estimated 15.4 tons fo nitrogen (pers. comm. D. Eckenrod, USEPA).  Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11]. From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to reduce the likelihood of an uncontrolled spill. @Switzer11 reported minimal impacts to nekton communities, although an increase in macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* was observed as a potential indication of nutrient eutrophication. Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on holding capacity of storage ponds [@Hu03].  Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.

@@

Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.  Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa) as common primary producers observed throughout the growing season and species associated with harmful algal blooms (HABs) as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.  Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* and *Pyrodinium bahamense* that can occur in the bay depending on salinity and temperature conditions during the growing season. Occurrence of both these species has historically been spatially distinct, with *K. brevis* originating in the Gulf oc Mexico and occuring in higher salinity portions of the bay, whereas *P. bahamense* has been observed consistently each year since 2008 in Old Tampa Bay (northwest segment) during the summer.  Data for *K. brevis* were also obtained from event-based monitoring samples collected by FWC and available from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).  Because of the increased occurrence of red tide samples in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St. Petersberg) impacted by the event. Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).

@@

A general narrative of events in Tampa Bay following release of wastewater from Piney Point on March 30th is shown in Figure \@ref(fig:timeline).  These events are supported by data summaries and results of statistical tests described below.  After the the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L. The inition diatom bloom did not perist past April.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the mouth of Tampa Bay and reached bloom concentrations by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, increased observations of *Lyngbya sp.* (a macroalgal cyanobacteria)

@@

```{r stktab}
cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and end-of-pipe samples from April 2021 for relevant water quality variables.  Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.  Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by the Environmental Protection Commission of Hillsborough County.  Effluent concentrations were taken from two samples on April 6th and 13th at the outflow at Port Manatee.  Averages were taken when two measured values were available from both effluent sample dates.  Missing values were not measured in the effluent.'
knitr::kable(stktab, booktabs = T, caption = cap.val)
```

```{r wqsumtab}
cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point. Variables are grouped by major areas interest for evaluating status and trends shown in Figure \\@ref(fig:map)a. Summaries are median, maximum, and minimum values. Total observations (N obs.) and values in rnage, above, or below normal ranges are also shown. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'
knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
```

>>>>

The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.  Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.  Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.  Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers. comm. D. Eckenrod, USEPA).  Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11]. From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to reduce the likelihood of an uncontrolled spill. @Switzer11 reported minimal impacts to nekton communities, although an increase in macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* was observed as a potential indication of nutrient eutrophication. Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on holding capacity of storage ponds [@Hu03].  Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.

@@

Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.  Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa) as common primary producers observed throughout the growing season and species associated with harmful algal blooms (HABs) as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.  Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* and *Pyrodinium bahamense* that can occur in the bay depending on salinity and temperature conditions during the growing season. Occurrence of both these species has historically been spatially distinct, with *K. brevis* originating in the Gulf of Mexico and occurring in higher salinity portions of the bay, whereas *P. bahamense* has been observed consistently each year since 2008 in Old Tampa Bay (northwest segment) during the summer.  Data for *K. brevis* were also obtained from event-based monitoring samples collected by FWC and available from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).  Because of the increased occurrence of red tide samples in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St. Petersberg) impacted by the event. Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).

@@

## Timeline of events from April 2021

A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline). After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Lyngbya sp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Lyngbya* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay by August.  The remainder of this section is quantitative description these events.

## Water quality response

```{r}
nwq <- sum(wqsumtab[['N obs.']])
outrng1 <- wqsumtab %>% 
  filter(`Water quality variable` == 'Secchi (m)') %>% 
  pull(below) %>%  
  sum
outrng2 <- wqsumtab %>% 
  filter(`Water quality variable` != 'Secchi (m)') %>% 
  pull(above) %>%  
  sum
outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 1)
```

From April to August 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).  Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others). The percent of observations outside of the normal range varied by location and parameter.  For chl-a, 51\% of the observations were above the normal range for area 1, whereas only 9\% and 24\% were above for areas 2 and 3, respectively.  Similarly, total nitrogen concentrations were above the normal range for 40\% of observations in area 1, whereas concentrations were above for only 12\% and  23\% of observations in areas 2 and 3.  Secchi observations for the period of observation were below the normal range for 46\% of observations in area 1 and for 24\% and 44\% of observations in areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., 50\% were above in area 1, 48\% in area 2).  Physical parameters (salinity, temperature) were generally within range over the five month period. Inorganic nitrogen (ammonia, nitrate/nitrite) were generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the effluent measurements in Table \@ref(tab:stktab). Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

Boxplots of total nitrogen, chl-a, and secchi depth show the temporal progression of observations by week and area relative to the normal ranges from long-term monitoring data (Figure \@ref(fig:wqtrnds)).  For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges beginning the week of April 4th and lasting through the month.  Concentrations remained similar to baseline conditions until June and July when median values were often the baseline range. Secchi observations in area 1 were below baseline ranges in April and June.  Observations in areas 2 and 3 were more often within the normal range, with some exceptions for total nitrogen and chl-a in area 3 during weeks of May, June, and July.  Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqtrnds). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1, total nitrogen and chl-a for area 2, and chl-a and secchi depth for area 3 (Table \@ref(tab:wqcmptab)). Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.  Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances.

Multivariate assessments of all water quality parameters that combined results across weeks by area showed significant associations between groups of variables (Figure \@ref(fig:pcacors)). The first principal component explained 41\%, 74\%, and 47\% of the variation among water quality parameters for areas 1, 2, and 3, respectively.  These components generally described a positive association between nutrients and chlorophyll concentrations and a negative association with secchi depth and dissolved oxygen.  The second and third principal components explained less of the variation among water quality parameters, such that 36\%, 20\%, and 34\% was explained by both axes for areas 1, 2, and 3, respectively.  The patterns explained by these axes were less clear, although physical drivers were partially explained in some cases (e.g., component 2 for area 2 explained a positive association for salinity and temperature, Figure \@ref(fig:pcacors)b).  Spearman rank correlations between the pairs of variables used for each of the PCAs showed significant associations.  Total nitrogen was positively correlated with chlorophyll and negatively associated with secchi depth for areas 1 and 2 ($p < 0.005$ and $p < 0.05$, respectively).  Dissolved oxygen was negatively associated with chl-a only in area 3 ($p < 0.005$), although the temporal progression likely changed over time in the other areas (e.g., increase followed by decrease).  Temperature generally had a positive association with chlorophyll ($p < 0.05$ for area 2, $p < 0.005$ for area 3).  pH was positively associated with chl-a in areas 1 and 2 ($p < 0.05$ for both).

## Seagrass and macroalgae trends

## Red tide and fish kill reports

@@

```{r stktab}
cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and end-of-pipe samples from April 2021 for relevant water quality variables.  Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.  Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by the Environmental Protection Commission of Hillsborough County.  Effluent concentrations were taken from two samples on April 6th and 13th at the outflow at Port Manatee.  Averages were taken when two measured values were available from both effluent sample dates.  Missing values were not measured in the effluent or stack water.'
knitr::kable(stktab, booktabs = T, caption = cap.val)
```

```{r wqsumtab}
cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point. Variables are grouped by major areas interest for evaluating status and trends shown in Figure \\@ref(fig:map)a. Summaries are median, maximum, and minimum values. Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'
knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
```
